Data Specialization for a Staged Lambda Calculus

### 2015-07-06 paper/semantics.tex
various small edits

One of the central concepts in the formalization of \lang\ is the notion of a \emph{world},
which includes and subsumes the concept of a stage.
Our system has three worlds: \bbonem, \bbonep, and \bbtwo.
Whereas there is only one world, \bbtwo, for second-stage computations,
there are two worlds corresponding to the first stage:
\bbonem\ for \emph{mixed} first-stage computations, which may contain second-stage subterms within $\next$ blocks,
and \bbonep\ for \emph{ground} first-stage computations, which may not.
The distinction between these two first-stage worlds is necessitated by our splitting algorithm
and will be discussed in \ref{sec:needGround}.

The abstract syntax of \lang\ is presented as a grammar in \ref{fig:grammar}.  
Note that some production rules are parameterized by world.
Our syntax also distinguishes values from general terms, 
with underline term constructor (``$\exv v$'') explicitly noting the inclusion of values into general terms.
There are three forms of value: \emph{partial values} for \bbonem, \emph{ground values} for \bbonep, and \emph{residuals} for \bbtwo.
All represent computations which have no remaining work in the first stage.

@@

\paragraph{World \bbonem.}
Like its ground counterpart, the \bbonem\ step judgement, $\stepmix e {e'}$, searches for redexes and reduces them in place.
It also descends into $\pure$ and $\next$ blocks, transitioning into one of the other kinds of steps there. 
The value form for \bbonem, called a partial value, comprises units, tuples, functions, $\pure$ blocks of ground values, and $\next$ blocks containing \emph{only a stage two variable}.
This strong requirement ensures that second-stage computations are not duplicated when partial values are substituted for variable.
This is a departure from the staged semantics of \cite{taha-thesis-99} and \cite{davies96}.
Whereas those semantics interpret values of type $\fut A$ to mean ``code of type $A$ that can be executed in the future''
ours interprets $\fut A$ to mean ``a reference to a value that will be accessible in the future.''
We find this interpretation to be more natural for our applications.
For example, consider:
\begin{lstlisting}
2`prev{`1`(fn x : $int => e') (next{`2`e`1`})`2`}`
\end{lstlisting}
We could reduce this to $\prev([\next~e/x]e')$, but this may potentially duplicate an
expensive computation $e$. Instead, we choose to \emph{hoist} $e$ outside, binding it
to a temporary variable $y$, and substituting that variable instead:
\begin{lstlisting} 
2`let val y = e in prev{`1`[next{`2`y`1`}/x]e'`2`}`
\end{lstlisting}
This behavior is implemented by the $\lift e {e'}$ judgement, called a \emph{hoisting step}.
These steps occur when a \next\ block has contents that are a residual but (to prevent loops) not when those contents are already a variable.
In essence, the rules for hoisting steps operate by ``bubbling up'' a subtitution to the innermost containing \prev,
where it is reified into a {\tt let} statement.

@@

The multistage function \texttt{qss} from \ref{sec:overview} has
type fitting the pattern \mbox{$A\to \fut (B \to C)$} at \bbonem.%
\footnote{We can rewrite \texttt{qss} in this curried form, or apply a
higher-order currying function.} 
To apply this function to arguments $\colmix a A$ and $\coltwo b B$, simply
evaluate the program:
\begin{lstlisting}
2`prev{`1`f a`2`} b`
\end{lstlisting}

We can also directly observe the reuse of first-stage computations across
multiple second-stage computations. To run many order statistics queries
$\texttt{k1},\dots,\texttt{km}$ on the same list, run the program:
\begin{lstlisting}
2`let val qs = prev{`1`qss `3`gr{`1`[7,4,2,5,9,...,3]`3`}`2`}`
2`in (qs k1,...,qs km)`
\end{lstlisting}
To see that the partial application of \texttt{qss} is only evaluated once,
observe that after the first stage of evaluation, the residual has the form:
\begin{lstlisting}
2`let val y = ...`
2`    val qs = y`
2`in (qs k1,...,qs km)`
\end{lstlisting}
where $\hat{y}$ is bound to a function containing no first-stage redexes.

>>>>

Although \lang\ describes computations that occur in two stages,
we find it helpful for the specification of splitting to differentiate terms into three \emph{worlds}.
A world is essentially a slightly finer classification than a stage.
Whereas there is only one world, \bbtwo, for second-stage computations,
there are two worlds corresponding to the first stage:
\bbonem\ for \emph{mixed} first-stage computations, which may contain second-stage subterms within $\next$ blocks,
and \bbonep\ for \emph{ground} first-stage computations, which may not.
The distinction between these two first-stage worlds is necessary for an efficient splitting algorithm
and will be discussed in \ref{sec:needGround}.

The abstract syntax of \lang\ is presented as a grammar in \ref{fig:grammar}.  
Note that some production rules are parameterized by world.
Our syntax also distinguishes values from general terms, 
with the underline term constructor (``$\exv v$'') explicitly noting the inclusion of values into general terms.
There are three forms of value: \emph{partial values} for \bbonem, \emph{ground values} for \bbonep, and \emph{residuals} for \bbtwo.
All represent computations which have no remaining work in the first stage.

@@

\paragraph{World \bbonem.}
Like its ground counterpart, the \bbonem\ step judgement, $\stepmix e {e'}$, searches for redexes and reduces them in place.
It also descends into $\pure$ and $\next$ blocks, transitioning into one of the other two steps ($\stepsym \bbonep$ or $\stepsym \bbtwo$) there. 
The value form for \bbonem, called a partial value, comprises units, tuples, functions, $\pure$ blocks of ground values, and $\next$ blocks containing \emph{only a stage two variable}.
This strong requirement ensures that second-stage computations are not duplicated when partial values are substituted for variable.
This is a departure from the staged semantics of \cite{taha-thesis-99} and \cite{davies96}.
Whereas those semantics interpret values of type $\fut A$ to mean ``code of type $A$ that can be executed in the future,''
ours interprets $\fut A$ to mean ``a reference to a value that will be accessible in the future.''
This difference is natural, given the goals of metaprogramming and partial evaluation versus our applications.
One consequence of the stronger requirement on partial values is that we will need a new kind of step to put terms into that form.
To illustrate, consider:
\begin{lstlisting}
2`prev{`1`(fn x : $int => e') (next{`2`e`1`})`2`}`
\end{lstlisting}
We could reduce this to $\prev([\next~e/x]e')$, but this may potentially duplicate an
expensive computation $e$. Instead, we choose to \emph{hoist} $e$ outside, binding it
to a temporary variable $y$, and substituting that variable instead:
\begin{lstlisting} 
2`let val y = e in prev{`1`[next{`2`y`1`}/x]e'`2`}`
\end{lstlisting}
This behavior is implemented by the $\lift e {e'}$ judgement, called a \emph{hoisting step}.
These steps occur when a \next\ block has contents that are a residual but (to prevent loops) not when those contents are already a variable.
In essence, the rules for hoisting steps operate by ``bubbling up'' a subtitution to the innermost containing \prev,
where it is reified into a {\tt let} statement.

@@

The multistage function \texttt{qss} from \ref{sec:overview} has a
type fitting the pattern \mbox{$A\to \fut (B \to C)$} at \bbonem.%
\footnote{We can rewrite \texttt{qss} in this curried form, or apply a
higher-order currying function.} 
To apply this function to arguments $\colmix a A$ and $\coltwo b B$, simply
evaluate the program:
\begin{lstlisting}
2`prev{`1`f a`2`} b`
\end{lstlisting}

We can also directly observe the reuse of first-stage computations across
multiple second-stage computations. To run many order statistics queries
$\texttt{k1},\dots,\texttt{km}$ on the same list, run the program:
\begin{lstlisting}
2`let val qs = prev{`1`qss `3`gr{`1`[7,4,2,5,9,...,3]`3`}`2`}`
2`in (qs k1,...,qs km)`
\end{lstlisting}
To see that the partial application of \texttt{qss} is only evaluated once,
observe that after the first stage of evaluation, the residual has the form:
\begin{lstlisting}
2`let val y = ...`
2`    val qs = y`
2`in (qs k1,...,qs km)`
\end{lstlisting}
where $y$ is bound to a function containing no first-stage redexes.

### 2015-07-06 paper/splitting.tex
various small edits

For a two-stage language like \lang, this means translating a multistage
object into a pair of monostage objects
which respectively embody the first-stage and second-stage content of the original.
Since \lang\ has three classes of multistage object, 
our splitting algorithm has three forms: 
\bbtwo-term splitting, \bbonem-term splitting, and partial value splitting.
These are given by the judgements $\splittwo e A p l r$, $\splitone e A c {l.r}$, and $v \vsplito \mval i q$.
The rules of all three judgements are given in \ref{fig:termSplit}.

@@

To provide intuition about the intent of partial value splitting, consider the
following partial value:
\begin{lstlisting}
1` (next{`2`yhat`1`}, (`3`gr{`1`injL 7`3`}`1`, next{`2`yhat`1`}))`
\end{lstlisting}
To construct the value $i$ representing its first-stage components,
splitting first redacts all second-stage (blue) parts, along with the
surrounding \texttt{next} annotations. 
The resulting ``holes'' in the term are replaced with
unit values.
%\begin{lstlisting}
%4`##########`1` (`4`######`1`, injL (gr{7}, `4`######`1`))`
%\end{lstlisting}
%The holes inside the term needs to be filled with something to convey the lack of information, so we use a unit values:
\begin{lstlisting}
1`((), (3`gr{`1`injL 7`3`}`,()))`
\end{lstlisting}
Finally, partial value splitting drops \texttt{gr} annotations yielding:
\begin{lstlisting}
1`((), (injL 7,()))`
\end{lstlisting}
To construct the residual $q$ (corresponding to second-stage computations) partial value splitting redacts all \texttt{gr} blocks
(replacing them with unit) and \texttt{next} annotations:
\begin{lstlisting}
2`(yhat, ((),yhat))`
\end{lstlisting}
A precise definition of the partial value splitting relation is given in
\ref{fig:valMask}.
Observe that partial value splitting does not lose any meaningful information (namely injection tags),
which is the validity requirement for a translation of value forms.

@@

This output is a multistage program in its own right.
It evaluates by reducing $p$ to $b$ in the first pass,
then plugging that result in for $l$ to produce $[b/l]r$,
which is evaluated in the second pass.
Since execution of the first pass serves to generate input for
the second pass, we say $p$ is a {\em precomputation} that produces a
{\em boundary value} ($b$) for the {\em resumer} ($l.r$).

We define the correctness of splitting 
by saying that $[b/l]r$, the \emph{applied resumer}, should be equivalent 
to the residual produced by direct evaluation, $q$.
We state this more precisely as \ref{thm:allCorrect},
where ``$\reduce e v$'' indicates standard monostage reduction of the term $e$ to the value $v$,
and ``$\equiv$'' indicates a monostage equivalence, which we cover later.

@@

All that remains is to prove \ref{lem:twoCorrect}.  
We will do this by induction on the derivation of the step.
However, since second-stage steps can depend on first-stage steps,
we first cover the definition of splitting terms at \bbonem.

@@

The correctness of \bbonem\ term splitting with respect to $\stepsym \bbonem$ steps is given in \ref{lem:oneCorrect}.
Like with the \bbtwo\ version, this requires that the applied resumer be equivalent both sides of the step.
But additionally, it also requires that first stage results be exactly identical.

The correctness of \bbonem\ term splitting with respect to hoisting steps is given in \ref{lem:hoistCorrect}.
Because hoisting steps are nothing but rearrangement of second stage code,
this lemma can use the strong requirement of idnetical combined terms.

@@

In this section, we prove the \ref{lem:oneCorrect} and \ref{lem:twoCorrect} lemmas
by considering a few exemplar cases.

@@

\paragraph{Compatility of {\tt pi1} at \bbtwo.}
We are given 
$\steptwo {\pio e} {\pio {e'}}$, 
$\splittwo {\pio e} A p l {\pio r}$, 
$\splittwo {\pio {e'}} A {p'} {l'} {\pio {r'}}$,
and $\reduce {p'} {b'}$.
Inversion of the first three yields
$\steptwo e {e'}$, $\splittwo e A p l r$, and $\splittwo {e'} A {p'} {l'} {r'}$.
Using \ref{lem:twoCorrect} inductively gives $\reduce p b$ and $[b/l]r \equiv [b'/l']r$.
From this, $[b/l]\pio r \equiv [b'/l']\pio {r}$ can be derived directly.
This pattern generalizes to all world \bbtwo\ compatibility rules.

\paragraph{Compatility of {\tt pi1} at \bbonem.}
We are given
$\stepmix {\pio e} {\pio {e'}}$, 
$\splitone {\pio e} A {\letin{\tup{y,z}}{c}{\tup{\pio {\exv y},\exv z}}} {l.\pio r}$, \\
\mbox{$\splitone {\pio {e'}} A {\letin{\tup{y,z}}{c'}{\tup{\pio {\exv y},\exv z}}} {l'.\pio {r'}}$},
and \\ $\reduce {\letin{\tup{y,z}}{c'}{\tup{\pio {\exv y},\exv z}}} {\tup{i,b'}}$.
Inversion of the first three yields 
$\stepmix e {e'}$, $\splitone e A c {l.r}$, and $\splitone {e'} A {c'} {l'.r'}$, 
and inversion of the reduction yields $\reduce {c'} {\tup{\tup{i,i_2},b'}}$.
Using \ref{lem:oneCorrect} inductively gives $\reduce c {\tup{\tup{i,i_2},b}}$ and $[b/l]r \equiv [b'/l']r$.
From this, we can derive $\reduce {\letin{\tup{y,z}}{c}{\tup{\pio {\exv y},\exv z}}} {\tup{i,b}}$ and 
$[b/l]\pio r \equiv [b'/l']\pio {r}$ directly.
This pattern generalizes to all world \bbonem\ compatibility rules.

@@

The splitting algorithm described in the previous subsections operates
purely on the local structure of \lang\ terms.  One artifact of this design
is that splitting \bbonem\ terms may generate resumers containing
unnecessary logic.  For example, the rule for splitting
\bbonem\ \texttt{case} terms inserts the tag from the \texttt{case}
argument into the boundary value, then decodes this tag in the
resumer. This logic occurs regardless of whether the terms forming the
branches of the \texttt{case} contain second-stage computations.
Worse, if this \texttt{case} appeared in the body of a recursive
function with no other second-stage computations, splitting would
generate a resumer with (useless) recursive calls.

Rather than attempt global optimization of the outputs of splitting,
we instead leverage the type system to indicate when a term contains
no second-stage computations by adding a third world \bbonep\ whose
terms are purely first-stage.  While not necessary for ordinary staged
term evaluation (\ref{sec:semantics}), this annotation is
particularly useful for splitting. Specifically, splitting $\pure e$
terms trivially yields a unit resumer.

@@

>>>>

For a two-stage language like \lang, this means translating a multistage
term into a pair of monostage terms
which respectively embody the first-stage and second-stage content of the original.
Since \lang\ has three classes of multistage term, 
our splitting algorithm has three forms: 
\bbtwo-term splitting, \bbonem-term splitting, and partial value splitting.
These are given by the judgements $\splittwo e A p l r$, $\splitone e A c {l.r}$, and $v \vsplito \mval i q$.
The rules of all three judgements are given in \ref{fig:termSplit}, 
and they operate locally on the structure of the term.
Note that there's no need to provide forms of splitting for ground terms or residuals,
since those classes of term are already monostage by construction.

@@

To provide intuition about the intent of partial value splitting, consider the
following partial value:
\begin{lstlisting}
1` (next{`2`yhat`1`}, (`3`gr{`1`injL 7`3`}`1`, next{`2`yhat`1`}))`
\end{lstlisting}
To construct the value $i$ representing its first-stage components,
splitting first redacts all second-stage (blue) parts, along with the
surrounding \texttt{next} annotations. 
The resulting ``holes'' in the term are replaced with
unit values.
%\begin{lstlisting}
%4`##########`1` (`4`######`1`, injL (gr{7}, `4`######`1`))`
%\end{lstlisting}
%The holes inside the term needs to be filled with something to convey the lack of information, so we use a unit values:
\begin{lstlisting}
1`((), (3`gr{`1`injL 7`3`}`,()))`
\end{lstlisting}
Finally, partial value splitting drops \texttt{gr} annotations yielding:
\begin{lstlisting}
1`((), (injL 7,()))`
\end{lstlisting}
To construct the residual $q$ (corresponding to second-stage computations) partial value splitting redacts all \texttt{gr} blocks
(replacing them with unit) and \texttt{next} annotations:
\begin{lstlisting}
2`(yhat, ((),yhat))`
\end{lstlisting}
A precise definition of the partial value splitting relation is given in
\ref{fig:termSplit}.
Observe that partial value splitting does not lose any meaningful information (namely injection tags),
which is the validity requirement for a translation of value forms.

@@

This output is a multistage program in its own right.
It evaluates by reducing $p$ to the value $b$ in the first pass,
then plugging that result in for $l$ to produce $[b/l]r$,
which is evaluated in the second pass.
Since execution of the first pass serves to generate input for
the second pass, we say $p$ is a {\em precomputation} that produces a
{\em boundary value} ($b$) for the {\em resumer} ($l.r$).

The validity of the splitting translation requires that this simple 
evaluate-and-plug semantics produces the same answer as the staged semantics of the previous section.
That is, $[b/l]r$ (the \emph{applied resumer}) should be equivalent 
to the residual produced by direct evaluation, $q$.
This condition, called \emph{correctness},
is stated more precisely as \ref{thm:allCorrect},
where ``$\reduce e v$'' indicates standard monostage reduction of the term $e$ to the value $v$,
and ``$\equiv$'' indicates a monostage equivalence, which we cover later.

@@

We prove \ref{lem:twoCorrect} by induction on the derivation of the step.
However, since second-stage steps can depend on first-stage steps,
we first cover the definition of splitting terms at \bbonem.

@@

The correctness of \bbonem\ term splitting with respect to $\stepsym \bbonem$ steps is given in \ref{lem:oneCorrect}.
As in the world \bbtwo\ version, correctness requires that the applied resumer from either side of the step be equivalent.
But additionally, it also requires that first stage results be exactly identical.

The correctness of \bbonem\ term splitting with respect to hoisting steps is given in \ref{lem:hoistCorrect}.
Because hoisting steps are nothing but rearrangement of second stage code,
this lemma can use the strong requirement of identical combined terms.

@@

In this section, we consider a few exemplar cases 
from the proofs of \ref{lem:oneCorrect} and \ref{lem:twoCorrect}.

@@

\paragraph{Compatibility of {\tt pi1} at \bbtwo.}
We are given 
$\steptwo {\pio e} {\pio {e'}}$, 
$\splittwo {\pio e} A p l {\pio r}$, 
$\splittwo {\pio {e'}} A {p'} {l'} {\pio {r'}}$,
and $\reduce {p'} {b'}$.
Inversion of the first three yields
$\steptwo e {e'}$, $\splittwo e A p l r$, and $\splittwo {e'} A {p'} {l'} {r'}$.
Using \ref{lem:twoCorrect} inductively gives $\reduce p b$ and $[b/l]r \equiv [b'/l']r$.
From this, $[b/l]\pio r \equiv [b'/l']\pio {r}$ can be derived directly.
This pattern generalizes to all world \bbtwo\ compatibility rules.

\paragraph{Compatibility of {\tt pi1} at \bbonem.}
We are given
$\stepmix {\pio e} {\pio {e'}}$, 
$\splitone {\pio e} A {\letin{\tup{y,z}}{c}{\tup{\pio {\exv y},\exv z}}} {l.\pio r}$, \\
\mbox{$\splitone {\pio {e'}} A {\letin{\tup{y,z}}{c'}{\tup{\pio {\exv y},\exv z}}} {l'.\pio {r'}}$},
and \\ $\reduce {\letin{\tup{y,z}}{c'}{\tup{\pio {\exv y},\exv z}}} {\tup{i,b'}}$.
Inversion of the first three yields 
$\stepmix e {e'}$, $\splitone e A c {l.r}$, and $\splitone {e'} A {c'} {l'.r'}$, 
and inversion of the reduction yields $\reduce {c'} {\tup{\tup{i,i_2},b'}}$.
Using \ref{lem:oneCorrect} inductively gives $\reduce c {\tup{\tup{i,i_2},b}}$ and $[b/l]r \equiv [b'/l']r$.
From this, we can derive $\reduce {\letin{\tup{y,z}}{c}{\tup{\pio {\exv y},\exv z}}} {\tup{i,b}}$ and 
$[b/l]\pio r \equiv [b'/l']\pio {r}$ directly.
This pattern generalizes to all world \bbonem\ compatibility rules.

@@

The splitting algorithm described in the previous subsections operates
purely on the local structure of \lang\ terms.  One artifact of this design
is that splitting \bbonem\ terms may generate resumers containing
unnecessary logic.  For example, the rule for splitting
\bbonem\ \texttt{caseg} terms inserts the tag from the \texttt{caseg}
argument into the boundary value, then decodes this tag in the
resumer. This logic occurs regardless of whether the terms forming the
branches of the \texttt{caseg} contain second-stage computations.
Worse, if this \texttt{caseg} appeared in the body of a recursive
function with no other second-stage computations, splitting would
generate a resumer with (useless) recursive calls.

Rather than attempt global optimization of the outputs of splitting,
we instead leverage the type system to indicate when a term contains
no second-stage computations by adding a third world \bbonep\ whose
terms are purely first-stage.  While not necessary for ordinary staged
term evaluation (\ref{sec:semantics}), this annotation is
particularly useful for splitting, since splitting $\pure e$
terms trivially yields a unit resumer.

@@

\TODO consider expanding, based on comments from reviews
